---
title: "Comparing Volleyball Stats by Class Year"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

We analyze whether collegiate experience level (Freshman, Sophomore, Junior, Senior) in NCAA Division 1 volleyball impacts player performance across multiple facets of the game: scoring, attacking efficiency, setting, serving, and defense.

```{r}
#| label: setup
library(tidyverse)
library(plotly)
library(DT)

volleyball <- read_csv("data/ncaa_d1_volleyball_2020_2026.csv", show_col_types = FALSE)

# Filter for standard class years and regular rotation players (>= 20 sets played)
vb_clean <- volleyball |>
  filter(Yr %in% c("Fr", "So", "Jr", "Sr"), S >= 20) |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman", "Sophomore", "Junior", "Senior")),
    kills_s = round(Kills / S, 2),
    assists_s = round(Assists / S, 2),
    aces_s = round(Aces / S, 2),
    digs_s = round(Digs / S, 2),
    blocks_s = round(coalesce(TB, coalesce(`Block Solos`, 0) + coalesce(`Block Assists`, 0)) / S, 2),
    points_s = round(PTS / S, 2),
    hit_pct = round(`Hit Pct`, 3)
  )
```

## Category Averages Across Class Years

Below are the per-set averages and attacking efficiencies across all four class standings from the 2020–2026 seasons. Click any column header to sort.

```{r}
#| label: class-summary
class_summary <- vb_clean |>
  group_by(`Class Standing` = class_year) |>
  summarize(
    `Players` = n(),
    `Points / Set` = round(mean(points_s, na.rm = TRUE), 2),
    `Kills / Set` = round(mean(kills_s, na.rm = TRUE), 2),
    `Hitting Pct` = round(mean(hit_pct, na.rm = TRUE), 3),
    `Digs / Set` = round(mean(digs_s, na.rm = TRUE), 2),
    `Blocks / Set` = round(mean(blocks_s, na.rm = TRUE), 2),
    `Aces / Set` = round(mean(aces_s, na.rm = TRUE), 2),
    `Assists / Set` = round(mean(assists_s, na.rm = TRUE), 2)
  )

datatable(class_summary, options = list(dom = 't', paging = FALSE), rownames = FALSE)
```

## Comparing Key Metrics by Class Year

This interactive chart compares scoring (Points), attacking (Kills), and defending (Digs and Blocks) per set across experience levels.

```{r}
#| label: multi-metric-chart
metrics_long <- class_summary |>
  select(`Class Standing`, `Points / Set`, `Kills / Set`, `Digs / Set`, `Blocks / Set`) |>
  pivot_longer(cols = -`Class Standing`, names_to = "Metric", values_to = "Average")

p_bars <- ggplot(metrics_long, aes(x = `Class Standing`, y = Average, fill = `Class Standing`)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Per-Set Production by Experience Level",
    x = NULL,
    y = "Average Per Set"
  ) +
  theme_minimal()

ggplotly(p_bars)
```

## Distribution of Total Points per Set

Hover over the boxplot below to examine the median, quartiles, and range of total points produced per set by class year.

```{r}
#| label: points-boxplot
p_box <- ggplot(vb_clean, aes(x = class_year, y = points_s, fill = class_year)) +
  geom_boxplot(outlier.alpha = 0.2, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  coord_cartesian(ylim = c(0, 6)) +
  labs(
    title = "Points per Set Distribution",
    x = "Class Standing",
    y = "Points per Set"
  ) +
  theme_minimal()

ggplotly(p_box)
```

## Interactive Player Comparison (Head-to-Head by Class Year)

Compare standout hitters **Sarah Franklin** and **Harper Murray** across each year of their college careers (**Freshman**, **Sophomore**, **Junior**, and **Senior**), or search the player directory below to build any custom head-to-head matchup.

```{=html}
<div id="h2h-comparison" class="card shadow mb-4 border-primary">
  <div class="card-header bg-primary text-white d-flex justify-content-between align-items-center flex-wrap gap-2">
    <div class="d-flex align-items-center gap-2">
      <span class="fw-bold fs-5">Head-to-Head Comparison</span>
      <span id="h2h-mode-badge" class="badge bg-light text-primary">Featured Career Track</span>
    </div>
    <div class="d-flex align-items-center gap-2 flex-wrap">
      <div class="btn-group btn-group-sm" role="group" aria-label="Select Year to Compare">
        <button type="button" class="btn btn-light active fw-semibold" id="btn-yr-Fr" onclick="setComparisonYear('Fr')">Freshman</button>
        <button type="button" class="btn btn-outline-light fw-semibold" id="btn-yr-So" onclick="setComparisonYear('So')">Sophomore</button>
        <button type="button" class="btn btn-outline-light fw-semibold" id="btn-yr-Jr" onclick="setComparisonYear('Jr')">Junior</button>
        <button type="button" class="btn btn-outline-light fw-semibold" id="btn-yr-Sr" onclick="setComparisonYear('Sr')">Senior</button>
      </div>
      <button type="button" class="btn btn-sm btn-outline-light d-none" id="btn-reset-featured" onclick="resetToFeatured()">
        ↺ Reset to Featured
      </button>
    </div>
  </div>
  <div class="card-body">
    <div class="row align-items-center mb-3">
      <div class="col-md-5 text-center p-3 rounded bg-light border" id="card-p1">
        <span class="badge bg-primary mb-1">Player 1</span>
        <h4 id="p1-name" class="text-primary mb-1">Sarah Franklin</h4>
        <p id="p1-meta" class="text-muted mb-0 small">Michigan St. (Big Ten) | Freshman | OH | 2020-2021</p>
      </div>
      <div class="col-md-2 text-center my-2 my-md-0">
        <span class="badge bg-danger fs-5 px-3 py-2 shadow-sm">VS</span>
        <div class="small text-muted mt-1 fw-bold" id="comparison-yr-label">Freshman Season</div>
      </div>
      <div class="col-md-5 text-center p-3 rounded bg-light border" id="card-p2">
        <span class="badge bg-success mb-1">Player 2</span>
        <h4 id="p2-name" class="text-success mb-1">Harper Murray</h4>
        <p id="p2-meta" class="text-muted mb-0 small">Nebraska (Big Ten) | Freshman | OH | 2023-2024</p>
      </div>
    </div>
    
    <div class="table-responsive">
      <table class="table table-sm table-striped table-hover text-center align-middle mb-0" id="h2h-table">
        <thead class="table-dark">
          <tr>
            <th class="text-start" style="width: 35%;">Stat Category</th>
            <th id="h2h-th-p1" style="width: 25%;">Sarah Franklin</th>
            <th id="h2h-th-p2" style="width: 25%;">Harper Murray</th>
            <th style="width: 15%;">Leader</th>
          </tr>
        </thead>
        <tbody id="h2h-tbody">
        </tbody>
      </table>
    </div>
  </div>
</div>

<script>
window.yearMap = { Fr: "Freshman", So: "Sophomore", Jr: "Junior", Sr: "Senior" };
window.abbrMap = { "Freshman": "Fr", "Sophomore": "So", "Junior": "Jr", "Senior": "Sr" };

window.franklinYears = {
  Fr: ["2020-2021", "Sarah Franklin", "Michigan St.", "Big Ten", "Freshman", "OH", 26, 4.13, 3.81, 0.162, 2.19, 0.17, 0.15, 0.08],
  So: ["2021-2022", "Sarah Franklin", "Michigan St.", "Big Ten", "Sophomore", "OH", 102, 4.50, 3.96, 0.224, 2.25, 0.58, 0.25, 0.26],
  Jr: ["2023-2024", "Sarah Franklin", "Wisconsin", "Big Ten", "Junior", "OH", 117, 4.62, 4.15, 0.300, 1.93, 0.53, 0.18, 0.13],
  Sr: ["2024-2025", "Sarah Franklin", "Wisconsin", "Big Ten", "Senior", "OH", 120, 4.96, 4.46, 0.289, 2.28, 0.49, 0.25, 0.15]
};

window.murrayYears = {
  Fr: ["2023-2024", "Harper Murray", "Nebraska", "Big Ten", "Freshman", "OH", 121, 3.82, 3.23, 0.237, 2.07, 0.54, 0.30, 0.37],
  So: ["2024-2025", "Harper Murray", "Nebraska", "Big Ten", "Sophomore", "OH", 121, 4.01, 3.40, 0.257, 2.43, 0.54, 0.32, 0.35],
  Jr: ["2025-2026", "Harper Murray", "Nebraska", "Big Ten", "Junior", "OH", 107, 4.21, 3.54, 0.295, 2.16, 0.60, 0.32, 0.33],
  Sr: null
};

window.p1PlayerName = "Sarah Franklin";
window.p2PlayerName = "Harper Murray";
window.currentComparisonYear = "Fr";
window.activeSlot = 'p1';
window.playerSeasonLookup = {};

function getPlayerData(playerName, yrAbbr) {
  if (!playerName) return null;
  var fullClass = window.yearMap[yrAbbr] || yrAbbr;
  var key = playerName.toLowerCase() + '|' + fullClass;
  if (window.playerSeasonLookup && window.playerSeasonLookup[key]) {
    return window.playerSeasonLookup[key];
  }
  if (playerName === "Sarah Franklin" && window.franklinYears && window.franklinYears[yrAbbr]) {
    return window.franklinYears[yrAbbr];
  }
  if (playerName === "Harper Murray" && window.murrayYears && window.murrayYears[yrAbbr]) {
    return window.murrayYears[yrAbbr];
  }
  return null;
}

function setComparisonYear(yr) {
  window.currentComparisonYear = yr;
  
  const years = ["Fr", "So", "Jr", "Sr"];
  years.forEach(y => {
    const btn = document.getElementById("btn-yr-" + y);
    if (btn) {
      if (y === yr) {
        btn.classList.remove("btn-outline-light");
        btn.classList.add("btn-light", "active");
      } else {
        btn.classList.remove("btn-light", "active");
        btn.classList.add("btn-outline-light");
      }
    }
  });

  const yrLabel = document.getElementById("comparison-yr-label");
  if (yrLabel) {
    yrLabel.innerText = (window.yearMap[yr] || yr) + " Season";
  }

  window.p1Data = getPlayerData(window.p1PlayerName, yr);
  window.p2Data = getPlayerData(window.p2PlayerName, yr);

  updateH2H();
}

function resetToFeatured() {
  window.p1PlayerName = "Sarah Franklin";
  window.p2PlayerName = "Harper Murray";
  const resetBtn = document.getElementById("btn-reset-featured");
  if (resetBtn) resetBtn.classList.add("d-none");
  const modeBadge = document.getElementById("h2h-mode-badge");
  if (modeBadge) {
    modeBadge.className = "badge bg-light text-primary";
    modeBadge.innerText = "Featured Career Track";
  }
  setComparisonYear('Fr');
}

function updateH2H() {
  var p1El = document.getElementById('p1-name');
  if (!p1El) return;
  
  var yrName = window.yearMap[window.currentComparisonYear] || window.currentComparisonYear;
  var p1DisplayName = window.p1Data ? window.p1Data[1] : window.p1PlayerName;
  var p2DisplayName = window.p2Data ? window.p2Data[1] : window.p2PlayerName;
  
  p1El.innerText = p1DisplayName;
  document.getElementById('h2h-th-p1').innerText = p1DisplayName;
  if (window.p1Data) {
    document.getElementById('p1-meta').innerText = window.p1Data[2] + ' (' + window.p1Data[3] + ') | ' + window.p1Data[4] + ' | ' + window.p1Data[5] + ' | ' + window.p1Data[0];
  } else {
    document.getElementById('p1-meta').innerText = "No recorded " + yrName + " season in dataset";
  }
  
  var p2El = document.getElementById('p2-name');
  p2El.innerText = p2DisplayName;
  document.getElementById('h2h-th-p2').innerText = p2DisplayName;
  if (window.p2Data) {
    document.getElementById('p2-meta').innerText = window.p2Data[2] + ' (' + window.p2Data[3] + ') | ' + window.p2Data[4] + ' | ' + window.p2Data[5] + ' | ' + window.p2Data[0];
  } else {
    document.getElementById('p2-meta').innerText = "No recorded " + yrName + " season in dataset";
  }
  
  const metrics = [
    { label: "Points / Set", p1: window.p1Data ? parseFloat(window.p1Data[7]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[7]) : NaN, digits: 2 },
    { label: "Kills / Set", p1: window.p1Data ? parseFloat(window.p1Data[8]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[8]) : NaN, digits: 2 },
    { label: "Hitting Percentage", p1: window.p1Data ? parseFloat(window.p1Data[9]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[9]) : NaN, digits: 3 },
    { label: "Digs / Set", p1: window.p1Data ? parseFloat(window.p1Data[10]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[10]) : NaN, digits: 2 },
    { label: "Blocks / Set", p1: window.p1Data ? parseFloat(window.p1Data[11]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[11]) : NaN, digits: 2 },
    { label: "Aces / Set", p1: window.p1Data ? parseFloat(window.p1Data[12]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[12]) : NaN, digits: 2 },
    { label: "Assists / Set", p1: window.p1Data ? parseFloat(window.p1Data[13]) : NaN, p2: window.p2Data ? parseFloat(window.p2Data[13]) : NaN, digits: 2 },
    { label: "Sets Played", p1: window.p1Data ? parseInt(window.p1Data[6]) : NaN, p2: window.p2Data ? parseInt(window.p2Data[6]) : NaN, digits: 0 }
  ];
  
  let rowsHtml = "";
  metrics.forEach(m => {
    let v1 = isNaN(m.p1) ? "—" : (m.digits === 3 ? m.p1.toFixed(3) : (m.digits === 0 ? m.p1 : m.p1.toFixed(2)));
    let v2 = isNaN(m.p2) ? "—" : (m.digits === 3 ? m.p2.toFixed(3) : (m.digits === 0 ? m.p2 : m.p2.toFixed(2)));
    let leader = "—";
    let p1Class = "";
    let p2Class = "";
    
    if (!isNaN(m.p1) && !isNaN(m.p2)) {
      if (m.p1 > m.p2) {
        leader = p1DisplayName;
        p1Class = "fw-bold text-primary";
      } else if (m.p2 > m.p1) {
        leader = p2DisplayName;
        p2Class = "fw-bold text-success";
      } else {
        leader = "Tie";
      }
    } else if (!isNaN(m.p1) && isNaN(m.p2)) {
      leader = p1DisplayName;
      p1Class = "fw-bold text-primary";
    } else if (isNaN(m.p1) && !isNaN(m.p2)) {
      leader = p2DisplayName;
      p2Class = "fw-bold text-success";
    }
    
    rowsHtml += `<tr>
      <td class="text-start fw-semibold">${m.label}</td>
      <td class="${p1Class}">${v1}</td>
      <td class="${p2Class}">${v2}</td>
      <td><span class="badge ${leader === p1DisplayName ? 'bg-primary' : (leader === p2DisplayName ? 'bg-success' : 'bg-secondary')}">${leader}</span></td>
    </tr>`;
  });
  
  document.getElementById('h2h-tbody').innerHTML = rowsHtml;
}

window.setActiveSlot = function(slot) {
  window.activeSlot = slot;
  const btn1 = document.getElementById("slot-btn-p1");
  const btn2 = document.getElementById("slot-btn-p2");
  const badge = document.getElementById("selection-status-badge");
  
  if (slot === 'p1') {
    if (btn1) { btn1.className = "btn btn-primary active fw-semibold"; }
    if (btn2) { btn2.className = "btn btn-outline-success fw-semibold"; }
    if (badge) {
      badge.className = "badge bg-primary text-wrap p-2";
      badge.innerText = "Ready: Click any row below to set Player 1";
    }
  } else {
    if (btn1) { btn1.className = "btn btn-outline-primary fw-semibold"; }
    if (btn2) { btn2.className = "btn btn-success active fw-semibold"; }
    if (badge) {
      badge.className = "badge bg-success text-wrap p-2";
      badge.innerText = "Ready: Click any row below to set Player 2";
    }
  }
};

window.handleTableSelection = function(d) {
  if (!d || d.length === 0) return;
  const pName = d[1];
  const pClass = d[4];
  const yrAbbr = window.abbrMap[pClass] || "Fr";
  const badge = document.getElementById("selection-status-badge");
  const resetBtn = document.getElementById("btn-reset-featured");
  if (resetBtn) resetBtn.classList.remove("d-none");

  const modeBadge = document.getElementById("h2h-mode-badge");
  if (modeBadge) {
    modeBadge.className = "badge bg-warning text-dark";
    modeBadge.innerText = "Custom Matchup";
  }

  if (window.activeSlot === 'p1') {
    window.p1PlayerName = pName;
    if (badge) {
      badge.innerText = "Selected " + pName + " (" + pClass + ") as Player 1! Next click will set Player 2.";
    }
    window.setActiveSlot('p2');
  } else {
    window.p2PlayerName = pName;
    if (badge) {
      badge.innerText = "Selected " + pName + " (" + pClass + ") as Player 2! Next click will set Player 1.";
    }
    window.setActiveSlot('p1');
  }

  setComparisonYear(yrAbbr);
};

if (document.readyState === "loading") {
  document.addEventListener("DOMContentLoaded", function() { setComparisonYear('Fr'); });
} else {
  setComparisonYear('Fr');
}
</script>
```

```{r}
#| label: prepare-player-data
player_table <- vb_clean |>
  select(
    Season, Player, Team, Conference, class_year, Pos, S,
    points_s, kills_s, hit_pct, digs_s, blocks_s, aces_s, assists_s
  ) |>
  rename(
    Class = class_year,
    Sets = S,
    `Pts/Set` = points_s,
    `Kills/Set` = kills_s,
    `Hit Pct` = hit_pct,
    `Digs/Set` = digs_s,
    `Blk/Set` = blocks_s,
    `Aces/Set` = aces_s,
    `Ast/Set` = assists_s
  )
```

## Player Directory (Custom Matchup Selector)

Search for any player by name, school, conference, or position. Click a row to assign that player to the comparison card above.

```{=html}
<div class="card p-3 mb-3 bg-light border shadow-sm">
  <div class="d-flex align-items-center justify-content-between flex-wrap gap-2">
    <div class="d-flex align-items-center gap-2">
      <span class="fw-bold">Next table click sets:</span>
      <div class="btn-group" role="group" id="targetSlotGroup">
        <button type="button" class="btn btn-primary active fw-semibold" id="slot-btn-p1" onclick="setActiveSlot('p1')">
          👤 Player 1 (Blue)
        </button>
        <button type="button" class="btn btn-outline-success fw-semibold" id="slot-btn-p2" onclick="setActiveSlot('p2')">
          👤 Player 2 (Green)
        </button>
      </div>
    </div>
    <div id="selection-status-badge" class="badge bg-primary text-wrap p-2">
      Ready: Click any row below to set Player 1
    </div>
  </div>
</div>
```

```{r}
#| label: player-directory-table
datatable(
  player_table,
  elementId = "player_directory_table",
  filter = "top",
  options = list(pageLength = 8, autoWidth = TRUE),
  rownames = FALSE,
  callback = JS("
    function(table) {
      window.playerSeasonLookup = {};
      var allData = table.data();
      for (var i = 0; i < allData.length; i++) {
        var r = allData[i];
        var key = r[1].toLowerCase() + '|' + r[4];
        if (!window.playerSeasonLookup[key] || (r[6] > window.playerSeasonLookup[key][6])) {
          window.playerSeasonLookup[key] = r;
        }
      }
      
      table.on('click', 'tbody tr', function() {
        var d = table.row(this).data();
        if (d && window.handleTableSelection) {
          window.handleTableSelection(d);
        }
      });
      
      if (window.setComparisonYear) {
        window.setComparisonYear(window.currentComparisonYear || 'Fr');
      }
    }
  ")
)
```

## Key Takeaways

1. **Overall Scoring**: Average points per set increases by **39%** from Freshman year (1.28) to Senior year (1.78).
2. **Attacking Volume & Efficiency**: Seniors average 1.39 kills per set with a .185 average hitting percentage, compared to 0.98 kills per set and .141 hitting percentage for freshmen.
3. **Net Defense**: Average blocks per set climb from 0.28 (freshmen) to 0.37 (seniors), a **32%** increase.
4. **Floor Defense**: Digs per set increase steadily each year, rising from 1.20 to 1.53 per set.
