---
title: "Comparing Volleyball Stats by Class Year"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

We analyze whether collegiate experience level (Freshman, Sophomore, Junior, Senior) in NCAA Division 1 volleyball impacts player performance across multiple facets of the game: scoring, attacking efficiency, setting, serving, and defense.

```{r}
#| label: setup
library(tidyverse)
library(plotly)
library(DT)

volleyball <- read_csv("data/ncaa_d1_volleyball_2020_2026.csv", show_col_types = FALSE)

# Filter for standard class years and regular rotation players (>= 20 sets played)
vb_clean <- volleyball |>
  filter(Yr %in% c("Fr", "So", "Jr", "Sr"), S >= 20) |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman", "Sophomore", "Junior", "Senior")),
    kills_s = round(Kills / S, 2),
    assists_s = round(Assists / S, 2),
    aces_s = round(Aces / S, 2),
    digs_s = round(Digs / S, 2),
    blocks_s = round(coalesce(TB, coalesce(`Block Solos`, 0) + coalesce(`Block Assists`, 0)) / S, 2),
    points_s = round(PTS / S, 2),
    hit_pct = round(`Hit Pct`, 3)
  )
```

## Category Averages Across Class Years

Below are the per-set averages and attacking efficiencies across all four class standings from the 2020–2026 seasons. Click any column header to sort.

```{r}
#| label: class-summary
class_summary <- vb_clean |>
  group_by(`Class Standing` = class_year) |>
  summarize(
    `Players` = n(),
    `Points / Set` = round(mean(points_s, na.rm = TRUE), 2),
    `Kills / Set` = round(mean(kills_s, na.rm = TRUE), 2),
    `Hitting Pct` = round(mean(hit_pct, na.rm = TRUE), 3),
    `Digs / Set` = round(mean(digs_s, na.rm = TRUE), 2),
    `Blocks / Set` = round(mean(blocks_s, na.rm = TRUE), 2),
    `Aces / Set` = round(mean(aces_s, na.rm = TRUE), 2),
    `Assists / Set` = round(mean(assists_s, na.rm = TRUE), 2)
  )

datatable(class_summary, options = list(dom = 't', paging = FALSE), rownames = FALSE)
```

## Comparing Key Metrics by Class Year

This interactive chart compares scoring (Points), attacking (Kills), and defending (Digs and Blocks) per set across experience levels.

```{r}
#| label: multi-metric-chart
metrics_long <- class_summary |>
  select(`Class Standing`, `Points / Set`, `Kills / Set`, `Digs / Set`, `Blocks / Set`) |>
  pivot_longer(cols = -`Class Standing`, names_to = "Metric", values_to = "Average")

p_bars <- ggplot(metrics_long, aes(x = `Class Standing`, y = Average, fill = `Class Standing`)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Per-Set Production by Experience Level",
    x = NULL,
    y = "Average Per Set"
  ) +
  theme_minimal()

ggplotly(p_bars)
```

## Distribution of Total Points per Set

Hover over the boxplot below to examine the median, quartiles, and range of total points produced per set by class year.

```{r}
#| label: points-boxplot
p_box <- ggplot(vb_clean, aes(x = class_year, y = points_s, fill = class_year)) +
  geom_boxplot(outlier.alpha = 0.2, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  coord_cartesian(ylim = c(0, 6)) +
  labs(
    title = "Points per Set Distribution",
    x = "Class Standing",
    y = "Points per Set"
  ) +
  theme_minimal()

ggplotly(p_box)
```

## Interactive Player Comparison (Head-to-Head)

Compare any two players side-by-side. Use the search bars in **Player 1** and **Player 2** below, then click any row in either table to update the comparison card instantly.

```{=html}
<div id="h2h-comparison" class="card shadow mb-4 border-primary">
  <div class="card-header bg-primary text-white d-flex justify-content-between align-items-center flex-wrap">
    <span class="fw-bold fs-5">Head-to-Head Player Comparison</span>
    <small class="badge bg-light text-primary">Click any row in Player 1 or Player 2 table below to update</small>
  </div>
  <div class="card-body">
    <div class="row align-items-center mb-3">
      <div class="col-md-5 text-center p-3 rounded bg-light border" id="card-p1">
        <span class="badge bg-primary mb-1">Player 1</span>
        <h4 id="p1-name" class="text-primary mb-1">Sarah Franklin</h4>
        <p id="p1-meta" class="text-muted mb-0 small">Wisconsin (Big Ten) | Junior | OH | 2023-2024</p>
      </div>
      <div class="col-md-2 text-center my-2 my-md-0">
        <span class="badge bg-danger fs-5 px-3 py-2 shadow-sm">VS</span>
      </div>
      <div class="col-md-5 text-center p-3 rounded bg-light border" id="card-p2">
        <span class="badge bg-success mb-1">Player 2</span>
        <h4 id="p2-name" class="text-success mb-1">Harper Murray</h4>
        <p id="p2-meta" class="text-muted mb-0 small">Nebraska (Big Ten) | Freshman | OH | 2023-2024</p>
      </div>
    </div>
    
    <div class="table-responsive">
      <table class="table table-sm table-striped table-hover text-center align-middle mb-0" id="h2h-table">
        <thead class="table-dark">
          <tr>
            <th class="text-start" style="width: 35%;">Stat Category</th>
            <th id="h2h-th-p1" style="width: 25%;">Sarah Franklin</th>
            <th id="h2h-th-p2" style="width: 25%;">Harper Murray</th>
            <th style="width: 15%;">Leader</th>
          </tr>
        </thead>
        <tbody id="h2h-tbody">
        </tbody>
      </table>
    </div>
  </div>
</div>

<script>
window.p1Data = ["2023-2024", "Sarah Franklin", "Wisconsin", "Big Ten", "Junior", "OH", 117, 4.62, 4.15, 0.300, 1.93, 0.53, 0.18, 0.13];
window.p2Data = ["2023-2024", "Harper Murray", "Nebraska", "Big Ten", "Freshman", "OH", 121, 3.82, 3.23, 0.237, 2.07, 0.54, 0.30, 0.37];

function updateH2H() {
  if (!window.p1Data || !window.p2Data) return;
  
  var p1El = document.getElementById('p1-name');
  if (!p1El) return;
  
  p1El.innerText = window.p1Data[1];
  document.getElementById('p1-meta').innerText = window.p1Data[2] + ' (' + window.p1Data[3] + ') | ' + window.p1Data[4] + ' | ' + window.p1Data[5] + ' | ' + window.p1Data[0];
  document.getElementById('h2h-th-p1').innerText = window.p1Data[1];
  
  document.getElementById('p2-name').innerText = window.p2Data[1];
  document.getElementById('p2-meta').innerText = window.p2Data[2] + ' (' + window.p2Data[3] + ') | ' + window.p2Data[4] + ' | ' + window.p2Data[5] + ' | ' + window.p2Data[0];
  document.getElementById('h2h-th-p2').innerText = window.p2Data[1];
  
  const metrics = [
    { label: "Points / Set", p1: parseFloat(window.p1Data[7]), p2: parseFloat(window.p2Data[7]), digits: 2 },
    { label: "Kills / Set", p1: parseFloat(window.p1Data[8]), p2: parseFloat(window.p2Data[8]), digits: 2 },
    { label: "Hitting Percentage", p1: parseFloat(window.p1Data[9]), p2: parseFloat(window.p2Data[9]), digits: 3 },
    { label: "Digs / Set", p1: parseFloat(window.p1Data[10]), p2: parseFloat(window.p2Data[10]), digits: 2 },
    { label: "Blocks / Set", p1: parseFloat(window.p1Data[11]), p2: parseFloat(window.p2Data[11]), digits: 2 },
    { label: "Aces / Set", p1: parseFloat(window.p1Data[12]), p2: parseFloat(window.p2Data[12]), digits: 2 },
    { label: "Assists / Set", p1: parseFloat(window.p1Data[13]), p2: parseFloat(window.p2Data[13]), digits: 2 },
    { label: "Sets Played", p1: parseInt(window.p1Data[6]), p2: parseInt(window.p2Data[6]), digits: 0 }
  ];
  
  let rowsHtml = "";
  metrics.forEach(m => {
    let v1 = isNaN(m.p1) ? "N/A" : (m.digits === 3 ? m.p1.toFixed(3) : (m.digits === 0 ? m.p1 : m.p1.toFixed(2)));
    let v2 = isNaN(m.p2) ? "N/A" : (m.digits === 3 ? m.p2.toFixed(3) : (m.digits === 0 ? m.p2 : m.p2.toFixed(2)));
    let leader = "—";
    let p1Class = "";
    let p2Class = "";
    
    if (!isNaN(m.p1) && !isNaN(m.p2)) {
      if (m.p1 > m.p2) {
        leader = window.p1Data[1];
        p1Class = "fw-bold text-primary";
      } else if (m.p2 > m.p1) {
        leader = window.p2Data[1];
        p2Class = "fw-bold text-success";
      } else {
        leader = "Tie";
      }
    }
    
    rowsHtml += `<tr>
      <td class="text-start fw-semibold">${m.label}</td>
      <td class="${p1Class}">${v1}</td>
      <td class="${p2Class}">${v2}</td>
      <td><span class="badge ${leader === window.p1Data[1] ? 'bg-primary' : (leader === window.p2Data[1] ? 'bg-success' : 'bg-secondary')}">${leader}</span></td>
    </tr>`;
  });
  
  document.getElementById('h2h-tbody').innerHTML = rowsHtml;
}

window.selectPlayer1 = function(row) {
  window.p1Data = row;
  updateH2H();
};

window.selectPlayer2 = function(row) {
  window.p2Data = row;
  updateH2H();
};

if (document.readyState === "loading") {
  document.addEventListener("DOMContentLoaded", updateH2H);
} else {
  updateH2H();
}
setTimeout(updateH2H, 300);
setTimeout(updateH2H, 1000);
</script>
```

```{r}
#| label: prepare-player-data
player_table <- vb_clean |>
  select(
    Season, Player, Team, Conference, class_year, Pos, S,
    points_s, kills_s, hit_pct, digs_s, blocks_s, aces_s, assists_s
  ) |>
  rename(
    Class = class_year,
    Sets = S,
    `Pts/Set` = points_s,
    `Kills/Set` = kills_s,
    `Hit Pct` = hit_pct,
    `Digs/Set` = digs_s,
    `Blk/Set` = blocks_s,
    `Aces/Set` = aces_s,
    `Ast/Set` = assists_s
  )
```

### Search Player 1

Use the search bar or column filters below. Click any row to assign this player to **Player 1**.

```{r}
#| label: player-table-1
datatable(
  player_table,
  elementId = "player_table_1",
  filter = "top",
  options = list(pageLength = 5, autoWidth = TRUE),
  rownames = FALSE,
  callback = JS("
    table.on('click', 'tr', function() {
      var d = table.row(this).data();
      if (d && window.selectPlayer1) {
        window.selectPlayer1(d);
      }
    });
  ")
)
```

### Search Player 2

Use the search bar or column filters below. Click any row to assign this player to **Player 2**.

```{r}
#| label: player-table-2
datatable(
  player_table,
  elementId = "player_table_2",
  filter = "top",
  options = list(pageLength = 5, autoWidth = TRUE),
  rownames = FALSE,
  callback = JS("
    table.on('click', 'tr', function() {
      var d = table.row(this).data();
      if (d && window.selectPlayer2) {
        window.selectPlayer2(d);
      }
    });
  ")
)
```

## Key Takeaways

1. **Overall Scoring**: Average points per set increases by **39%** from Freshman year (1.28) to Senior year (1.78).
2. **Attacking Volume & Efficiency**: Seniors average 1.39 kills per set with a .185 average hitting percentage, compared to 0.98 kills per set and .141 hitting percentage for freshmen.
3. **Net Defense**: Average blocks per set climb from 0.28 (freshmen) to 0.37 (seniors), a **32%** increase.
4. **Floor Defense**: Digs per set increase steadily each year, rising from 1.20 to 1.53 per set.
