---
title: "Comparing Volleyball Stats by Class Year"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

We analyze whether collegiate experience level (Freshman, Sophomore, Junior, Senior) in NCAA Division 1 volleyball impacts player performance across multiple facets of the game: scoring, attacking efficiency, setting, serving, and defense.

```{r}
#| label: setup
library(tidyverse)
library(plotly)
library(DT)

volleyball <- read_csv("data/ncaa_d1_volleyball_2020_2026.csv", show_col_types = FALSE)

# Filter for standard class years and regular rotation players (>= 20 sets played)
vb_clean <- volleyball |>
  filter(Yr %in% c("Fr", "So", "Jr", "Sr"), S >= 20) |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman", "Sophomore", "Junior", "Senior")),
    kills_s = round(Kills / S, 2),
    assists_s = round(Assists / S, 2),
    aces_s = round(Aces / S, 2),
    digs_s = round(Digs / S, 2),
    blocks_s = round(coalesce(TB, coalesce(`Block Solos`, 0) + coalesce(`Block Assists`, 0)) / S, 2),
    points_s = round(PTS / S, 2),
    hit_pct = round(`Hit Pct`, 3)
  )
```

## Category Averages Across Class Years

Below are the per-set averages and attacking efficiencies across all four class standings from the 2020–2026 seasons. Click any column header to sort.

```{r}
#| label: class-summary
class_summary <- vb_clean |>
  group_by(`Class Standing` = class_year) |>
  summarize(
    `Players` = n(),
    `Points / Set` = round(mean(points_s, na.rm = TRUE), 2),
    `Kills / Set` = round(mean(kills_s, na.rm = TRUE), 2),
    `Hitting Pct` = round(mean(hit_pct, na.rm = TRUE), 3),
    `Digs / Set` = round(mean(digs_s, na.rm = TRUE), 2),
    `Blocks / Set` = round(mean(blocks_s, na.rm = TRUE), 2),
    `Aces / Set` = round(mean(aces_s, na.rm = TRUE), 2),
    `Assists / Set` = round(mean(assists_s, na.rm = TRUE), 2)
  )

datatable(class_summary, options = list(dom = 't', paging = FALSE), rownames = FALSE)
```

## Comparing Key Metrics by Class Year

This interactive chart compares scoring (Points), attacking (Kills), and defending (Digs and Blocks) per set across experience levels.

```{r}
#| label: multi-metric-chart
metrics_long <- class_summary |>
  select(`Class Standing`, `Points / Set`, `Kills / Set`, `Digs / Set`, `Blocks / Set`) |>
  pivot_longer(cols = -`Class Standing`, names_to = "Metric", values_to = "Average")

p_bars <- ggplot(metrics_long, aes(x = `Class Standing`, y = Average, fill = `Class Standing`)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Per-Set Production by Experience Level",
    x = NULL,
    y = "Average Per Set"
  ) +
  theme_minimal()

ggplotly(p_bars)
```

## Distribution of Total Points per Set

Hover over the boxplot below to examine the median, quartiles, and range of total points produced per set by class year.

```{r}
#| label: points-boxplot
p_box <- ggplot(vb_clean, aes(x = class_year, y = points_s, fill = class_year)) +
  geom_boxplot(outlier.alpha = 0.2, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  coord_cartesian(ylim = c(0, 6)) +
  labs(
    title = "Points per Set Distribution",
    x = "Class Standing",
    y = "Points per Set"
  ) +
  theme_minimal()

ggplotly(p_box)
```

## Interactive Player Explorer

Search, filter by conference, position, or team, and sort players across all statistical categories.

```{r}
#| label: player-table
player_table <- vb_clean |>
  select(
    Season, Player, Team, Conference, class_year, Pos, S,
    points_s, kills_s, hit_pct, digs_s, blocks_s, aces_s, assists_s
  ) |>
  rename(
    Class = class_year,
    Sets = S,
    `Pts/Set` = points_s,
    `Kills/Set` = kills_s,
    `Hit Pct` = hit_pct,
    `Digs/Set` = digs_s,
    `Blk/Set` = blocks_s,
    `Aces/Set` = aces_s,
    `Ast/Set` = assists_s
  )

datatable(
  player_table,
  filter = "top",
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE
)
```

## Key Takeaways

1. **Overall Scoring**: Average points per set increases by **39%** from Freshman year (1.28) to Senior year (1.78).
2. **Attacking Volume & Efficiency**: Seniors average 1.39 kills per set with a .185 average hitting percentage, compared to 0.98 kills per set and .141 hitting percentage for freshmen.
3. **Net Defense**: Average blocks per set climb from 0.28 (freshmen) to 0.37 (seniors), a **32%** increase.
4. **Floor Defense**: Digs per set increase steadily each year, rising from 1.20 to 1.53 per set.
