---
title: "The Veteran Advantage: Learning to Win at the Net"
format: html
execute:
  echo: false
  warning: false
  message: false
---

When evaluating collegiate volleyball attackers, raw kill totals often grab headlines. But in high-level NCAA Division 1 play, rallies frequently hinge on attacking discipline: putting the ball to the floor without handing free points to the opponent on unforced errors. To understand how collegiate hitters mature, we examined individual season statistics from over 15,000 player-seasons across six NCAA seasons (2020–2026), filtering for active attackers with at least 100 attack attempts.

```{r}
#| label: veteran-advantage-plot
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

library(tidyverse)

volleyball <- read_csv("data/ncaa_d1_volleyball_2020_2026.csv", show_col_types = FALSE)

trends <- volleyball |>
  filter(Yr %in% c("Fr", "So", "Jr", "Sr"), `Total Attacks` >= 100) |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman", "Sophomore", "Junior", "Senior"))
  ) |>
  group_by(class_year) |>
  summarize(
    `Kill Rate` = mean(Kills / `Total Attacks`, na.rm = TRUE) * 100,
    `Error Rate` = mean(Errors / `Total Attacks`, na.rm = TRUE) * 100
  ) |>
  pivot_longer(
    cols = c(`Kill Rate`, `Error Rate`),
    names_to = "Metric",
    values_to = "Percentage"
  )

ggplot(trends, aes(x = class_year, y = Percentage, color = Metric, group = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = ifelse(trends$Metric == "Kill Rate", -1.2, 1.8),
    fontface = "bold",
    size = 3.8,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("Kill Rate" = "#1b7837", "Error Rate" = "#d73027")) +
  scale_y_continuous(limits = c(10, 42), labels = \(x) paste0(x, "%")) +
  labs(
    title = "The Veteran Shift: Rising Kills and Declining Errors",
    subtitle = "Attack outcome rates across NCAA Division 1 hitters (2020–2026, min. 100 attacks)",
    x = "Class Standing",
    y = "Percentage of Total Attacks",
    color = NULL,
    caption = "Source: NCAA Division 1 Volleyball Statistics"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
```

The data reveals a steady, two-sided evolution. As attackers progress from their freshman to senior campaigns, their kill rate climbs monotonically from 34.5% to 36.9%. At the same time, their attack error rate declines each year, dropping from 16.4% down to 14.3%. Together, these compounding adjustments explain why seniors hit at an average efficiency of .226 compared to .181 for freshmen: veterans generate more offensive points while donating fewer errors.

## Position Breakdown: Outside Hitters vs. Middle Blockers

Not all hitting roles are created equal. Middle Blockers (MB) attack quick sets close to the setter, often facing single blockers, while Outside Hitters (OH) carry the burden of out-of-system "bailout" sets against formed double blocks. Does collegiate experience help both roles equally?

```{r}
#| label: position-comparison-plot
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

pos_trends <- volleyball |>
  filter(Yr %in% c("Fr", "So", "Jr", "Sr"), `Total Attacks` >= 100, Pos %in% c("OH", "MB")) |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman", "Sophomore", "Junior", "Senior")),
    position_full = if_else(Pos == "OH", "Outside Hitters (OH)", "Middle Blockers (MB)")
  ) |>
  group_by(position_full, class_year) |>
  summarize(
    `Kill Rate` = mean(Kills / `Total Attacks`, na.rm = TRUE) * 100,
    `Error Rate` = mean(Errors / `Total Attacks`, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(`Kill Rate`, `Error Rate`),
    names_to = "Metric",
    values_to = "Percentage"
  )

ggplot(pos_trends, aes(x = class_year, y = Percentage, color = Metric, group = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = ifelse(pos_trends$Metric == "Kill Rate", -1.2, 1.8),
    fontface = "bold",
    size = 3.3,
    show.legend = FALSE
  ) +
  facet_wrap(~position_full) +
  scale_color_manual(values = c("Kill Rate" = "#1b7837", "Error Rate" = "#d73027")) +
  scale_y_continuous(limits = c(10, 48), labels = \(x) paste0(x, "%")) +
  labs(
    title = "Veteran Improvement Across Playing Positions",
    subtitle = "Comparing Outside Hitters (heavy out-of-system load) and Middle Blockers (quick attacks)",
    x = "Class Standing",
    y = "Percentage of Total Attacks",
    color = NULL,
    caption = "Source: NCAA Division 1 Volleyball Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```

The split shows that experience pays off differently depending on the role. Middle Blockers maintain high kill rates from day one (39.5% as freshmen) because of their physical height and set speed, yet they still trim errors down from 15.5% to 13.7%. Outside Hitters experience an even more noticeable transformation: their error rate drops by more than two full points (17.4% to 15.2%) while kill rates jump from 32.3% to 34.0%, demonstrating how seasoned pins learn to manage difficult balls.

## Case Study: Marta Levinska (Arizona State)

Aggregate statistics show broad patterns, but individual careers bring the trend to life. Marta Levinska arrived at Arizona State in 2020 as a high-potential international opposite/outside hitter. Over four consecutive seasons, her trajectory tracked the veteran curve with textbook precision.

```{r}
#| label: player-case-study
#| fig-width: 8
#| fig-height: 4.2
#| fig-align: center

marta <- volleyball |>
  filter(Player == "Marta Levinska") |>
  mutate(
    class_year = factor(Yr, levels = c("Fr", "So", "Jr", "Sr"),
                        labels = c("Freshman\n(2020-21)", "Sophomore\n(2021-22)", "Junior\n(2022-23)", "Senior\n(2023-24)")),
    `Kill Rate` = (Kills / `Total Attacks`) * 100,
    `Error Rate` = (Errors / `Total Attacks`) * 100
  ) |>
  select(class_year, `Kill Rate`, `Error Rate`) |>
  pivot_longer(cols = c(`Kill Rate`, `Error Rate`), names_to = "Metric", values_to = "Percentage")

ggplot(marta, aes(x = class_year, y = Percentage, color = Metric, group = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = ifelse(marta$Metric == "Kill Rate", -1.2, 1.8),
    fontface = "bold",
    size = 3.6,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("Kill Rate" = "#1b7837", "Error Rate" = "#d73027")) +
  scale_y_continuous(limits = c(5, 52), labels = \(x) paste0(x, "%")) +
  labs(
    title = "Four-Year Trajectory: Marta Levinska (Arizona State)",
    subtitle = "Kill rate rose from 33.1% to 45.8% as hitting percentage climbed from .204 to .348",
    x = "Season & Standing",
    y = "Percentage of Total Attacks",
    color = NULL,
    caption = "Source: NCAA Division 1 Volleyball Statistics"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
```

As a freshman, Levinska delivered a productive 245 kills with a .204 hitting percentage. By her senior campaign in 2023–24, she racked up 571 kills while pushing her kill rate to 45.8% and cutting her error rate to 11.0%—earning AVCA All-America honors and guiding Arizona State to the NCAA Sweet 16. Her growth exemplifies how collegiate attackers combine strength, shot selection, and court vision to dominate at the net.
